---
title: "Emergency Hospital Admissions in Aberdeen"
html_notebook: default
---

```{r setup, include=FALSE}
library(here)
library(tidyverse)
library(janitor)
library(sf)
library(mapview)
library(broom)
library(gt)
library(cowplot)
library(plotly)
theme_set(theme_cowplot()) 
```

```{r, include=FALSE}
#load data and shapefile

simd_zones <- read_csv(here("data", "SIMD+2020v2+-+datazone+lookup.csv")) %>%
  clean_names() %>%
  rename(dz_name = d_zname)

simd_indicators <- read_csv(here("data", "SIMD2020v2_indicators.csv")) %>%
  clean_names()

datazone_sf <- st_read(here("data", "sc_dz_11.shp"), quiet = T)

simd <- 
  left_join(simd_indicators, simd_zones, by = c("data_zone" = "dz"))

rm(simd_indicators, simd_zones)
```

```{r, include=FALSE}
simd <-
  simd %>%
  mutate(
    across(simd2020v2_rank:population, ~ as.integer(.)),
    emerg = (emerg/100))

simd_aberdeen <-
  simd %>%
  filter(council_area == "Aberdeen City") %>%
  arrange(simd2020v2_rank) %>%
  mutate(aberdeen_rank = 
           row_number(),
         aberdeen_decile =
           as_factor(ntile(.$simd2020v2_rank, 10)),
         aberdeen_vigintile =
           as_factor(ntile(.$simd2020v2_rank, 20)),
         aberdeen_percentile =
           as_factor(ntile(.$simd2020v2_rank, 100)))

#emerg_rank_model <- lm(emerg ~ aberdeen_rank, simd_aberdeen)
emerg_rank_model <- 
  lm(emerg ~ poly(aberdeen_rank, 2), simd_aberdeen)

simd_aberdeen<-
  simd_aberdeen %>%
  augment(emerg_rank_model, .) %>%
  mutate(outlier = case_when(
    .resid >  1*sd(.$.resid) ~ "higher than expected",
    .resid < -1*sd(.$.resid) ~ "lower than expected"))
```

Identifying areas that rely on **emergency medical care** helps prioritise where to focus **preventative care**.  

Here we look at emergency hospital admissions for each neighbourhood in Aberdeen. Admission rates are standarised for age and sex differences between neighbourhoods.  

Emergency hospital admissions in Aberdeen are **`r 100-(median(simd_aberdeen$emerg)*100)`% lower than the Scottish average**.  

However, there are **large inequalities in emergency admissions across the city**. In different neighbourhoods, the **emergency admissions rate varies from `r simd_aberdeen %>% arrange(emerg) %>% slice_tail() %>% .$emerg*100-100`% HIGHER than the Scottish average to `r 100 -(simd_aberdeen %>% arrange(emerg) %>% slice_head() %>% .$emerg*100)`% LOWER**. 

```{r, include=FALSE}
#select Aberdeen City datazones
#join to make separate shapefile

aberdeen_data_zones <- pull(simd_aberdeen, data_zone)

aberdeen_sf <- filter(datazone_sf, DataZone %in% aberdeen_data_zones)

aberdeen_sf <-
  left_join(aberdeen_sf, simd_aberdeen, by = c("DataZone" = "data_zone"))
```

<br>

### Mapping Emergency Hospital Admissions
High-emergency and low-emergency neighbourhoods cluster together. **Heathryfold, Middlefied, Mastrick and Northfield have much higher emergency admission rates** than expected given the age and sex breakdown of the population. **Deeside areas have much lower** emergency admission rates than expected.  
<br> 
Hover or click to see area names
```{r, echo=FALSE}
aberdeen_sf %>%
  select(DataZone, Name, emerg, geometry) %>%
  mapview(
    map.types = "OpenStreetMap",
    zcol = "emerg", 
    label = aberdeen_sf$Name,
    layer.name = "Emergency Admissions <br>vs Scottish Average of 1",
    alpha.regions = 0.75,
    na.label = ""
    )
```
<br>

### Deprived Areas Use More Emergency Care 
**Emergency admissions rise steeply as area deprivation rises.**  

This graph shows emergency admission rates for the 283 neighbourhoods in Aberdeen compared to their level of socioeconomic disadvantage (from the Scottish Government's Index of Multiple Deprivation).  

Hover or click to see area names
```{r, echo=FALSE}
q <-
simd_aberdeen %>%
  ggplot(aes(aberdeen_rank, emerg, label = dz_name)) +
  geom_hline(yintercept = 1, color = "dark grey", size = 1, alpha = 0.75) +
  annotate("text", x = 285, y = 1, label = "Scottish\naverage", size = 3) +
  geom_point(alpha = 0.5, colour = "dark red") +
  scale_x_reverse(breaks = c(1, 100, 200, 300), 
                  labels = c("1\nmost deprived", "100", "200", "300\nleast deprived"), expand = c(0, 30)) +
  scale_y_continuous(limits = c(0, 2)) +
  theme(legend.position = "NA",
        plot.margin = margin(1,1,1,1, unit = "cm")) +
  labs(x = "Deprivation Rank", 
       y = "Emergency Hosptial Admission Rate\n") +
  theme(axis.text = element_text(size = 8),
        axis.title = element_text(size = 12)) 

ggplotly(q, tooltip = "label") %>%
  layout(margin = list(
    r = 10, l = 1, t = 20, b = 1))
```
<br>

### Comparing Emergency Admissions in Similar Areas 
**Within neighbourhoods with equal deprivation, there can be quite different emergency admissions rates**.  

A neighbourhood in George Street has an emergency admission rate `r (1-(simd_aberdeen %>% filter(dz_name == "George Street - 04") %>% .$emerg))*100`% LOWER than the Scottish average. An area with the equal deprivation in Sheddocksley has an emergency admission rate `r (1-(simd_aberdeen %>% filter(dz_name == "Sheddocksley - 01") %>% .$emerg))*-100`% HIGHER than the Scottish average.  

**City centre & Kittybrewster neighbourhoods have lower than expected emergency admission rates given their deprivation**. Northfield & Mastrick areas have higher than expected rates, even given their high level of deprivation. Old Aberdeen and Torry have a mix.

```{r, echo=FALSE, warning=FALSE, message=FALSE}

aberdeen_sf %>% 
  select(DataZone, Name, emerg, outlier, geometry) %>%
  mapview(
    map.types = "OpenStreetMap",
    zcol = "outlier", 
    col.regions = c("red", "dark green"),
    label = aberdeen_sf$Name,
    layer.name = md("Emergency Hosptial Admissions,<br>(adjusted for deprivation)"),
    alpha.regions = 0.75,
    na.label = "as expected"
    )
```


<br>

### Data for all Aberdeen Areas
```{r, echo=FALSE}
simd_aberdeen %>%
  select(data_zone, dz_name, emerg, simd2020v2_decile) %>%
  arrange(desc(emerg)) %>%
  gt() %>%
  tab_header(
    title = "Emergency hospital admissions rate relative to Scottish average (age-sex adjusted)", 
    subtitle = "All Aberdeen data zones, sorted by highest emergency admissions") %>%
  cols_label(
    data_zone = "Data Zone",
    dz_name = "Name",
    emerg = "Emergency Admissions",
    simd2020v2_decile = "SIMD decile (lower is more deprived)") %>%
  tab_options(container.height = 400,
              data_row.padding = 1)
```
<br>
<br>
<br>

### Methods
This analysis uses emergency admissions to non-psychiatric and non-obstetric NHS hospitals for the four-year period of financial years 2014-2015 to 2017-2018. Data are given for the 283 Scottish Government Data Zones in Aberdeen city (each with a population of about 750). Admission rates are indirectly age-and-sex standardised and given relative to the expected value for Scotland.

Data source: Scottish Government Index of Multiple Deprivation (data_zone and emerg variables) (<https://www.gov.scot/collections/scottish-index-of-multiple-deprivation-2020/>)

Methodology: <https://www.gov.scot/binaries/content/documents/govscot/publications/statistics/2020/09/simd-2020-technical-notes/documents/simd-2020-technical-notes/simd-2020-technical-notes/govscot%3Adocument/SIMD%2B2020%2Btechnical%2Bnotes.pdf?forceDownload=true>  

Analysis code: <https://github.com/JessButler/emergency_admissions>

Author: Jess Butler [jessicabutler\@abdn.ac.uk](mailto:jessicabutler@abdn.ac.uk){.email}

```{r, include=FALSE}
simd_aberdeen %>%
  select(dz_name, emerg) %>%
  arrange(desc(emerg)) %>%
  slice(1:5, 279:283) %>%
  gt() %>%
  cols_label(
    dz_name = "",
    emerg = md("**Admission rate<br>compared to Scotland**")) %>%
  tab_options(container.height = 400,
              data_row.padding = 1,
              table.font.size = 12,
              table.align = "left") %>%
   tab_row_group(
    label = md("**Areas with Lowest Emergency Rate**"),
    rows = 6:10) %>%
  tab_row_group(
    label = md("**Areas with Highest Emergency Rate**"),
    rows = 1:5) 
```