---
title: "Emergency Hospital Admissions in Aberdeen"
html_notebook: default
---

```{r setup, include=FALSE}
library(here)
library(tidyverse)
library(janitor)
library(sf)
library(mapview)
library(gt)
library(cowplot)
library(plotly)
theme_set(theme_cowplot()) 
```

```{r, include=FALSE}
#load data and shapefile

simd_zones <- read_csv(here("data", "SIMD+2020v2+-+datazone+lookup.csv")) %>%
  clean_names() %>%
  rename(dz_name = d_zname)

simd_indicators <- read_csv(here("data", "SIMD2020v2_indicators.csv")) %>%
  clean_names()

datazone_sf <- st_read(here("data", "sc_dz_11.shp"), quiet = T)

simd <- 
  left_join(simd_indicators, simd_zones, by = c("data_zone" = "dz"))

rm(simd_indicators, simd_zones)
```

```{r, include=FALSE}
simd <-
  simd %>%
  mutate(
    across(simd2020v2_rank:population, ~ as.integer(.)),
    emerg = (emerg/100))

simd_aberdeen <-
  simd %>%
  filter(council_area == "Aberdeen City") %>%
  arrange(simd2020v2_rank) %>%
  mutate(aberdeen_rank = 
           row_number(),
         aberdeen_decile =
           as_factor(ntile(.$simd2020v2_rank, 10)),
         aberdeen_vigintile =
           as_factor(ntile(.$simd2020v2_rank, 20)),
         aberdeen_percentile =
           as_factor(ntile(.$simd2020v2_rank, 100)))
```

Identifying areas that rely on **emergency** healthcare helps prioritise where to focus **preventative** care. Below is an analysis of emergency hospital admissions across the city of Aberdeen.  

Emergency admissions from 2014-2017 are shown for the 283 neighbourhoods (each with a population of ~750). Admission rates are compared to the Scottish average (always standarised for age and sex characteristics of the neighbourhood).

Emergency hospital admissions in Aberdeen are **`r 100-(median(simd_aberdeen$emerg)*100)`% lower than the Scottish average**.  

However, there are **large inequalities in emergency admissions across the city**. In different neighbourhoods, **emergency admissions rate varies from `r simd_aberdeen %>% arrange(emerg) %>% slice_tail() %>% .$emerg*100-100`% higher than the Scottish average to `r 100 -(simd_aberdeen %>% arrange(emerg) %>% slice_head() %>% .$emerg*100)`% lower**. 

```{r, include=FALSE}
#select Aberdeen City datazones
#join to make separate shapefile

aberdeen_data_zones <- pull(simd_aberdeen, data_zone)

aberdeen_sf <- filter(datazone_sf, DataZone %in% aberdeen_data_zones)

aberdeen_sf <-
  left_join(aberdeen_sf, simd_aberdeen, by = c("DataZone" = "data_zone"))
```
<br>

### Mapping Emergency Hospital Admissions
Heathryfold, Middlefied, Mastrick and Northfield have **much higher** emergency admission rates than expected given the age and sex breakdown of the population. The Deeside areas have **much lower** emergency admission rates than expected.  
<br> 
Hover or click to see area names
```{r, echo=FALSE}
aberdeen_sf %>%
  select(DataZone, Name, emerg, geometry) %>%
  mapview(
    map.types = "OpenStreetMap",
    zcol = "emerg", 
    label = aberdeen_sf$Name,
    layer.name = "Emergency Admissions <br>vs Scottish Average",
    alpha.regions = 0.75,
    na.label = ""
    )
```
  
  
```{r, include=FALSE}
vigintile_values <-
  simd_aberdeen %>%
  group_by(aberdeen_vigintile) %>%
  summarise(median = median(emerg))
```
<br>

### Deprived Areas Need More Emergency Care 
**Emergency admissions rise steeply as area deprivation rises.**  

This graph shows emergency admission rates for the 283 neighbourhoods in Aberdeen compared to their level of socioeconomic disadvantage (from the Scottish Government's Index of Multiple Deprivation).  

Hover or click to see area names
```{r, echo=FALSE}
q <-
simd_aberdeen %>%
  ggplot(aes(aberdeen_rank, emerg, label = dz_name)) +
  geom_hline(yintercept = 1, color = "dark grey", size = 1, alpha = 0.5) +
  annotate("text", x = 50, y = 0.3, label = "Fewer emergencies\nthan expected", size = 3) +
  annotate("text", x = 200, y = 1.6, label = "More emergencies\nthan expected", size = 3) +
  annotate("text", x = 285, y = 1, label = "Scottish\naverage", size = 3) +
  geom_point(alpha = 0.5, colour = "dark red") +
  scale_x_reverse(breaks = c(1, 100, 200, 300), 
                  labels = c("1\nmost deprived", "100", "200", "300\nleast deprived"), expand = c(0, 30)) +
  scale_y_continuous(limits = c(0, 2)) +
  theme(legend.position = "NA",
        plot.margin = margin(1,1,1,1, unit = "cm")) +
  labs(x = "Deprivation Rank", 
       y = "Emergency Hosptial Admission Rate\n") +
  theme(axis.text = element_text(size = 8),
        axis.title = element_text(size = 12))

ggplotly(q, tooltip = "label") %>%
  layout(margin = list(
    r = 10, l = 1, t = 20, b = 1))
```
<br>

### Data for all Aberdeen Areas
```{r, echo=FALSE}
simd_aberdeen %>%
  select(data_zone, dz_name, emerg, simd2020v2_decile) %>%
  arrange(desc(emerg)) %>%
  gt() %>%
  tab_header(
    title = "Emergency hospital admissions rate relative to Scottish average (age-sex adjusted)", 
    subtitle = "All Aberdeen data zones, sorted by highest emergency admissions") %>%
  cols_label(
    data_zone = "Data Zone",
    dz_name = "Name",
    emerg = "Emergency Admissions",
    simd2020v2_decile = "SIMD decile (lower is more deprived)") %>%
  tab_options(container.height = 400,
              data_row.padding = 1)
```
<br>
<br>
<br>

### Methods
This analysis uses emergency admissions to non-psychiatric and non-obstetric NHS hospitals for the four-year period of 2014 to 2017. Data are given for the 283 Scottish Government Data Zones in Aberdeen city. Admission rates are indirectly age and sex standardised and given relative to the expected value for Scotland.

Data source: Scottish Government Index of Multiple Deprivation (data_zone and emerg variables) (<https://www.gov.scot/collections/scottish-index-of-multiple-deprivation-2020/>)

Methodology: <https://www.gov.scot/binaries/content/documents/govscot/publications/statistics/2020/09/simd-2020-technical-notes/documents/simd-2020-technical-notes/simd-2020-technical-notes/govscot%3Adocument/SIMD%2B2020%2Btechnical%2Bnotes.pdf?forceDownload=true>  

Analysis code: <https://github.com/JessButler/emergency_admissions>

Author: Jess Butler [jessicabutler\@abdn.ac.uk](mailto:jessicabutler@abdn.ac.uk){.email}