---
title:
html_notebook: default
---

## Inequalities in Emergency Hospital Admission in Aberdeen

```{r setup, include=FALSE}
library(here)
library(tidyverse)
library(janitor)
library(sf)
library(mapview)
library(gt)
library(cowplot)
library(plotly)
theme_set(theme_cowplot()) 
```

```{r, include=FALSE}
#load data and shapefile

simd_zones <- read_csv(here("data", "SIMD+2020v2+-+datazone+lookup.csv")) %>%
  clean_names() %>%
  rename(dz_name = d_zname)

simd_indicators <- read_csv(here("data", "SIMD2020v2_indicators.csv")) %>%
  clean_names()

datazone_sf <- st_read(here("data", "sc_dz_11.shp"), quiet = T)

simd <- 
  left_join(simd_indicators, simd_zones, by = c("data_zone" = "dz"))

rm(simd_indicators, simd_zones)
```

```{r, include=FALSE}
simd <-
  simd %>%
  mutate(
    across(simd2020v2_rank:population, ~ as.integer(.)))

simd_aberdeen <-
  simd %>%
  filter(council_area == "Aberdeen City") %>%
  arrange(simd2020v2_rank) %>%
  mutate(aberdeen_rank = 
           row_number(),
         aberdeen_decile =
           as_factor(ntile(.$simd2020v2_rank, 10)),
         aberdeen_vigintile =
           as_factor(ntile(.$simd2020v2_rank, 20)),
         aberdeen_percentile =
           as_factor(ntile(.$simd2020v2_rank, 100)))
```
  
Aberdeen's median emergency hospital admission rate is `r median(simd_aberdeen$emerg)` (**lower than Scotland's** overall value of 100).  

However, there are large differences in emergency admissions across the city. **`r simd_aberdeen %>% arrange(emerg) %>% slice_tail() %>% .$intermediate_zone`** (zone `r simd_aberdeen %>% arrange(emerg) %>% slice_tail() %>% .$data_zone`) has the highest admission rate in Aberdeen at `r simd_aberdeen %>% arrange(emerg) %>% slice_tail() %>% .$emerg`. This means emergency admissions in this area are **`r simd_aberdeen %>% arrange(emerg) %>% slice_tail() %>% .$emerg - 100`% higher** than expected given the age and sex breakdown of the population.


```{r, include=FALSE}
#select Aberdeen City datazones
#join to make separate shapefile

aberdeen_data_zones <- pull(simd_aberdeen, data_zone)

aberdeen_sf <- filter(datazone_sf, DataZone %in% aberdeen_data_zones)

aberdeen_sf <-
  left_join(aberdeen_sf, simd_aberdeen, by = c("DataZone" = "data_zone"))
```

### Mapping Emergency Hospital Admissions
Heathryfold, Middlefied, Mastrick and Northfield have much higher than expected emergency admission rates.  
Hover or click to see area names
```{r, echo=FALSE}
aberdeen_sf %>%
  mapview(
    map.types = "OpenStreetMap",
    zcol = "emerg", 
    label = aberdeen_sf$Name,
    layer.name = "Emergency Rate",
    alpha.regions = 0.75,
    na.label = ""
    )
```
  
  
```{r, include=FALSE}
vigintile_values <-
  simd_aberdeen %>%
  group_by(aberdeen_vigintile) %>%
  summarise(median = median(emerg))
```
  
### Emergency Hospital Admissions and Socioeconomic Deprivation 
**Emergency admissions rise as an area's deprivation rises.**  
Areas in the most deprived 5% of Aberdeen have a median emergency admission rate more than twice that of the least deprived 5%.  
Hover or click to see area names
```{r, echo=FALSE}
q <-
simd_aberdeen %>%
  ggplot(aes(aberdeen_rank, emerg, label = dz_name)) +
  geom_hline(yintercept = 100, color = "dark grey", size = 1, alpha = 0.5) +
  annotate("text", x = 10, y = 85, label = "Scotland\naverage", size = 3) +
  geom_point(alpha = 0.5, colour = "dark red") +
  scale_x_reverse(breaks = c(1, 100, 200, 300), 
                  labels = c("1\nmost deprived", "100", "200", "300\nleast deprived"), expand = c(0, 30)) +
  scale_y_continuous(limits = c(0, 200)) +
  theme(legend.position = "NA",
        plot.margin = margin(1,1,1,1, unit = "cm")) +
  labs(x = "Deprivation Rank", 
       y = "Emergency Hosptial Admissions") +
  theme(axis.text = element_text(size = 8),
        axis.title = element_text(size = 12))

ggplotly(q, tooltip = "label") 
```
  
  
### Data
```{r, echo=FALSE}
simd_aberdeen %>%
  select(data_zone, dz_name, emerg, simd2020v2_decile) %>%
  arrange(desc(emerg)) %>%
  gt() %>%
  tab_header(
    title = "Aberdeen emergency hospital admission rates", 
    subtitle = "All Aberdeen data zones, sorted by highest emergency admissions") %>%
  cols_label(
    data_zone = "Data Zone",
    dz_name = "Name",
    emerg = "Emergency Hospital Admissions",
    simd2020v2_decile = "SIMD decile (lower is more deprived)")  %>%
  tab_options(container.height = 400,
              data_row.padding = 1)
```
  
  
  
### Methods
This analysis uses emergency admissions to non-psychiatric and non-obstetric NHS hospitals for the four-year period of 2014 to 2017. Data are given for the 283 Scottish Government Data Zones in Aberdeen city. Admission rates are indirectly age and sex standardised and can be compared to the value of 100 for Scotland overall.

Data source: Scottish Government Index of Multiple Deprivation (<https://www.gov.scot/collections/scottish-index-of-multiple-deprivation-2020/>)

Analysis code: <https://github.com/JessButler/emergency_admissions>

Author: Jess Butler [jessicabutler\@abdn.ac.uk](mailto:jessicabutler@abdn.ac.uk){.email}